---
title: "Working with DHS sibling history data"
author: "Dennis M. Feehan with collaboration of Claude (Opus 5)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with DHS sibling history data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The Demographic and Health Surveys (DHS) collect sibling survival histories in an
optional maternal mortality module in the women's questionnaire.
`prep_dhs_sib_histories()` prepares those data for the estimators in this
package.

This vignette covers what DHS sibling data looks like, and what we learned
validating the package against published DHS estimates. Two companions:
"Preparing DHS data" walks through what the prep does step by step, and
"Working with MICS sibling history data" is the MICS equivalent of this one.

## Which DHS surveys can be used

The module is optional, so most DHS surveys do not have it. The practical test is
whether the individual recode (`IR`) file contains `mm`-prefixed columns:
`mm1_01`, `mm2_01` and so on. Surveys without the module simply have none.

The roster numbering has been stable since DHS phase 2, which makes life much
easier than it is with MICS. There are really only two variants:

| | DHS 2--5 | DHS 6--8 |
|---|---|---|
| Varmap | `sibhist_varmap_dhs2` … `_dhs5` | `sibhist_varmap_dhs6` … `_dhs8` |
| Roster items | `mm1`--`mm15` | `mm1`--`mm16` |
| Violence/accident item? | No | `mm16`, **from 2016 onwards** |
| Estimand | Pregnancy-related only | Maternal *and* pregnancy-related |

The catch is in that bolded qualifier. `mm16` was introduced in surveys fielded
from 2016, which does **not** line up with the phase boundary --- Rwanda 2014-15
is phase 7 and has no `mm16`. In a sample of 43 surveys with a sibling roster
spanning phases 2 through 8, only **5** carried `mm16` at all, so maternal
mortality proper is computable for a small minority and pregnancy-related
mortality for all of them.

Passing a varmap that names a column the data does not have is fine. The prep
records it and carries on:

```{r, eval=FALSE, purl=FALSE}
prep$summ$sib.cols.notfound
#> "mm16"
```

and `sib.maternal.death.date` comes back all `NA`, which is the honest answer.

## A minimal example

DHS publishes the roster **wide** --- one row per respondent, with a numbered
column per sibling per item --- so the prep reshapes it.

```{r, eval=FALSE, purl=FALSE}
library(siblingsurvival)
library(haven)

ir <- read_dta("RWIR61FL.DTA")

prepped <- prep_dhs_sib_histories(ir, varmap = sibhist_varmap_dhs6)

prepped$ego.dat   # one row per respondent
prepped$sib.dat   # one row per reported sibling
prepped$summ      # a one-row summary
```

Unlike MICS, the survey code is carried in the data (`v000`), so there is no
`survey` argument to supply.

To add the pregnancy-related and maternal death columns:

```{r, eval=FALSE, purl=FALSE}
prep_dhs_sib_histories(ir, varmap = sibhist_varmap_dhs6, add_maternal = TRUE)
```

## Special features of DHS files

**Weights are multiplied by 1,000,000.** `v005` is the women's weight scaled by
a million, so the prep defaults to `weight.scale = 1e6`. (MICS weights already
average 1, which is why `prep_mics_sib_histories()` defaults to `1` instead ---
passing the DHS default there would divide every weight by a million and nothing
downstream would complain.)

**One coded item carries the whole estimand.** Where MICS asks three separate
binaries, the DHS records everything in `mm9`:

| `mm9` | meaning | pregnancy-related | maternal |
|---|---|---|---|
| 2 | died while pregnant | yes | yes |
| 3 | died during delivery | yes | yes |
| 4 | since delivery --- never assigned in practice | yes | yes |
| 5 | within six weeks of a delivery | yes | yes |
| 6 | between six weeks and two months of a delivery | **yes** | no |

Code 6 is what makes pregnancy-related the **two-month** quantity and maternal
the **42-day** one.

**`mm16` is not asked when the death was during delivery.** For `mm9 = 3` it is
missing by design, not by non-response, so a missing `mm16` has to count as "not
an accident". Requiring `mm16 == 0` would silently drop every delivery death ---
in The Gambia 2019-20, all 40 in-window delivery deaths have `mm16` missing and
none of the `mm9 = 2` or `mm9 = 5` deaths do.

**Some files need a non-default encoding.** Gabon 2000 (`GAIR41FL`) fails
`haven::read_dta()` outright with *"Unable to convert string to the requested
encoding (invalid byte sequence)"*. `encoding = "latin1"` reads it. It is worth
checking whether a pipeline is silently skipping such a file rather than
erroring.

**A column can be present and entirely empty.** Burkina Faso 2003 has an `mm9`
column in which all 249,540 values are missing, so no pregnancy-related death can
be identified and the count is exactly zero --- which reads as a mortality
finding rather than as a survey that never coded the module.
`add_maternal_deaths()` warns about this.

**Sex codes other than 1 and 2 occur.** The DHS labels `mm1 = 8` as "don't know",
and some surveys carry an unlabelled 9 --- Gabon 2000 has 163 of them. These are
set to `NA` and dropped, and counted in `summ$miss.sex`.

## Validation against published estimates

The DHS Program publishes the code that produces its report tables:
[`DHS-Indicators-Stata`](https://github.com/DHSProgram/DHS-Indicators-Stata),
directory `Chap16_AM`. `AM_rates.do` is 1,376 lines by Thomas Pullum, and its
header states that it *"agrees exactly with DHS procedures, except for
confidence intervals"*. Transcribing it literally into R --- see
`data-raw/dhs-validation/stata-reference-replica.R` --- gives an independent
implementation to check against, and settles several conventions the reports
themselves leave undocumented.

### All-cause adult mortality

The cleanest test of the estimator, because it involves no cause recode at all.
Five surveys spanning DHS phases 4 to 8, each against its own published table:

| Survey | Table | deaths | exposure | age-adj. rate |
|---|---|---|---|---|
| Malawi 2000 | 12.2 | 1.0002 / 0.9999 | **1.000000** | 1.0012 / 0.9967 |
| Rwanda 2005 | 12.3 | 1.0007 / 0.9994 | **1.000000** | 1.0006 / 1.0004 |
| Rwanda 2014-15 | 16.3 | 0.9997 / 0.9994 | **1.000000** | 0.9999 / 1.0005 |
| Gambia 2019-20 | 14.1 | 0.9995 / 0.9997 | **1.000000** | 0.9989 / 1.0010 |
| Rwanda 2010 | 16.3 | 0.9991 / 0.9998 | **1.000000** | 0.975 / 1.070 |

Ratios of the package's value to the published one, women / men. **Exposure
matches to the person-year in every cell of every survey** --- 70 cells. Deaths
match to rounding. Rwanda 2010's rates are discussed below.

**Each report chooses its own window, and they differ.** Malawi 2000 and Gambia
publish 0--6 years before the survey; the three Rwanda surveys publish 0--4. The
window is part of the target, not a constant, and using one window for all of
them is the easiest way to "fail" a validation that would otherwise pass exactly.

`q15_to_50()` end-to-end on Gambia 2019-20 gives 113.5 and 124.4 against a
published 114 and 124.

### Age standardisation uses the respondents, for both sexes

This one is worth stating plainly because the reference code suggests otherwise.
`get_age_distributions` in `AM_rates.do:450` takes the men's age distribution
from the men's recode (`MR`) file. Published reports do not appear to have been
produced that way --- they standardise **both** sexes by the age distribution of
the survey respondents, which is what `get_ego_age_distn(only_females = TRUE)`
returns:

| Survey | respondents' distribution | `MR` distribution | published |
|---|---|---|---|
| Malawi 2000 | 11.064 | 11.162 | **11.1** |
| Rwanda 2005 | **7.393** | 7.285 | **7.39** |
| Rwanda 2014-15 | **2.961** | 2.881 | **2.96** |
| Gambia 2019-20 | **3.133** | --- | **3.13** |

So a published male adult mortality rate is reproducible from the women's file
alone. `get_ego_age_distn(only_females = FALSE)` warns if you ask it to split a
single-sex sample, and points at this.

### Pregnancy-related and maternal deaths

Rwanda 2010's Table 16.4 is a *pregnancy-related* table despite its heading ---
the report says so explicitly, defining the deaths as occurring in the window
"even if the death is due to causes that are not pregnancy related". Every cell
reproduces:

| Age | 15-19 | 20-24 | 25-29 | 30-34 | 35-39 | 40-44 | 45-49 | Total |
|---|---|---|---|---|---|---|---|---|
| Ours | 4.1 | 16.1 | 19.5 | 23.4 | 16.8 | 7.7 | 3.1 | 90.7 |
| Published | 4 | 16 | 20 | 23 | 17 | 8 | 3 | 91 |

The Gambia 2019-20's Table 14.3 *is* a maternal table --- 42-day window, "deaths
due to accidents or violence are excluded" --- and it reproduces exactly too: 65.0
against a published 65, with maternal deaths at 17.3% of female deaths against a
published 17.3%.

### The whole sample

All 43 surveys with a sibling roster have been run through the prep and compared
against the reference implementation
(`data-raw/dhs-validation/breadth-check.R`). Every one completes; female
exposure, female and male all-cause deaths and pregnancy-related deaths match the
reference **exactly** in all 43; no estimate cell is `NA`; and no survey has a
missing date of birth or weight after the prep.

Male exposure matches exactly in 42 of 43. Gabon 2012 is high by 6.1
person-years in 105,599 --- 0.006% --- from a single sibling whose `mm4` is
missing and whose date of birth the package recovers from his reported age,
where the reference drops him. Arguably the package is the better behaviour
there; `AM_rates.do` assumes `mm4` is always coded.

## Rwanda 2010's published summary rows are unreliable

Its age-specific cells reproduce exactly --- deaths, exposure, pregnancy-related
deaths, every age group, both sexes. Its **totals** do not, in three independent
ways:

* the age-adjusted rates match the *crude* rates (3.083 and 3.636, printed as 3.1
  and 3.6) rather than the standardised ones, despite an "age-adjusted" footnote;
* Table 16.4's total exposure prints 165,352 against age rows that sum to
  120,900 --- and 165,352 is exactly the *seven-year* exposure, so that one cell
  was filled from a different run than the rest of the table;
* the report's own maternal mortality ratio of 476 is consistent with the
  standardised rate, not with the rate printed beside it.

Use its age-specific cells; do not use its totals as a validation target.

## What will not agree with reports, by design

* **`asdr.ind`.** The individual-visibility estimator is this package's
  contribution and appears in no published table.
* **Confidence intervals.** `AM_rates.do` says its agreement with DHS holds
  "except for confidence intervals". DHS fits Poisson models and reports
  log-scale intervals; this package bootstraps and reports percentile intervals.
  Note also that DHS's own sampling-error appendix for maternal mortality comes
  from a *separate* CSPro application rather than from `AM_rates.do`.
* **Subpopulation rates.** The DHS Program explicitly declines to compute
  maternal mortality within subpopulations, so there is nothing to compare
  against.
* **The maternal mortality ratio.** MMR is the age-standardised rate divided by a
  general fertility rate from the women's birth history --- a module this package
  does not read. The replica computes it, for validation only.

## Pregnancy-related versus maternal

Two quantities, and they are not the same thing:

| | Window | Excludes violence/accident? | `mm9` codes | Available |
|---|---|---|---|---|
| Pregnancy-related | two months | no | 2--6 | every survey with a roster |
| Maternal | 42 days | yes (`mm16`) | 2--5 | only where `mm16` exists |

`add_maternal_deaths()` computes both, following `AM_rates.do:725` and `:728`.
Neither uses `mm12`, the time-since-delivery band: the reference states plainly
that "mm12 is not needed" and drops it before the roster is reshaped.

One further wrinkle, which limits maternal mortality more than the `mm16`
availability alone suggests. **The `mm9` 5/6 split is a questionnaire artefact,
not a clean partition.** Of the five surveys carrying `mm16`, four use both
codes, but South Africa 2016 uses only code 6 --- which cannot literally mean
that no woman died within six weeks of delivery, and much more likely means its
questionnaire asked only about the two-month window. Where that happens,
`mm9 <= 5` drops *every* postpartum death and the maternal count degenerates to
"died pregnant or during delivery".

Pregnancy-related mortality spans both codes and so is insensitive to this. It is
computable for every survey with a roster, which is a good reason to prefer it
when comparing across surveys.

## A note on the 2016 PRMR redefinition

`AM_rates.do:307-320` carries, *inside a comment block and therefore never
executed*, a rule introduced as "Important for redefinition of Pregnancy Related
Mortality Ratio (PRMR) in surveys from 2016 onwards":

```
If mm9=2, and mm16=1 or 2, recode mm9 to 1
```

That is: a death during pregnancy reported as violence or an accident stops
counting as pregnancy-related. The DHS Program documents the rule but the code it
ships does not apply it, so published figures do not reflect it either.
`add_maternal_deaths(prmr.accident.recode = TRUE)` applies it; the default is
`FALSE`, which is what reproduces published tables.
