---
title: "Estimating maternal mortality rates from sibling history data"
author: "Dennis M. Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating death rates from sibling history data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(siblingsurvival)
library(tidyverse)
# be sure you have at least version 0.0.2.9000 of surveybootstrap
# run devtools::install_github('dfeehan/surveybootstrap')
# for the most recent version
library(surveybootstrap)  

# this is helfpul for timing
library(tictoc)
```


## Overview

This vignette shows how to use the `siblingsurvival` package to estimate 
quantities related to maternal mortality. You might find it helpful to also
read the [vignette about estimating mortality rates](sibling-estimates.html); 
much of the material here is based on that vignette. In fact, the process for 
estimating maternal mortality rates is very similar to estimating all-cause 
death rates, with a few modifications which we will explain in this vignette.

It will be helpful to define a few important terms before we start our analysis:

* `ego` - an ego is a survey respondent
* `cell` - a cell is a generic group for which we wish to produce estimates.
 Usually, a cell is defined by a time period, an age range, and a sex. 
 So, for example, a cell might be women who were age 30-34 in 2015.

As an overview, we'll start from a survey dataset, such as
the one provided by the DHS. We then use `prep_dhs_sib_histories` to produce two datasets:

* one dataset has a row for each survey respondent/ego
* one dataset has a row for each sibling who is reported by a survey respondent

We'll then calculate estimates from the sibling histories in three main steps:

1. Create an **esc** dataset, so called because there is  row for each **e**go X **s**ibling X **c**ell
2. Aggregate this esc dataset up into an **ec** dataset, which has a row for the reports made by each ego for each cell
3. Aggregate this ec dataset up into estimates for death rates, using either the individual or aggregate visibility approach (or both)

### Opening up the demo datasets

We'll start by opening up the demonstration DHS dataset.

```{r}
data(model_dhs_dat)
```

### Using `prep_dhs_sib_histories` to get ego and sib datasets

Now we'll use `prep_dhs_sib_histories` to ready the sibling histories for analysis.
(Please see the [Preparing Data](preparing-data.html) vignette for more details.)
This is the step that will produce two datasets: one with a row for each
survey respondent/ego; and one with a row for each reported sibling.
Note that we include the parameter `add_maternal=TRUE` to indicate that we
want columns related to maternal/pregnancy-related mortality to be added.

```{r}
prepped <- prep_dhs_sib_histories(model_dhs_dat,
                                  varmap = sibhist_varmap_dhs6,
                                  add_maternal = TRUE,
                                  keep_missing = FALSE)

# we'll only keep the variable we will need for this analysis
ex.ego <- prepped$ego.dat %>%
  ## add a 'sex' variable
  mutate(sex = 'f') %>%
  select(caseid, 
         psu, 
         stratum_analysis, 
         stratum_design, 
         cluster, 
         age.cat, sex, wwgt)
ex.sib <- prepped$sib.dat
```
Let's take a look at the datasets we've produced. First, here's a dataset that has information about survey respondents. (We'll refer to these survey respondents as 'ego'):

```{r}
glimpse(ex.ego)
```

And here's a long-form version of the sibling history data -- there's one row for each reported sibling.

```{r}
glimpse(ex.sib)
```

### Laying the groundwork

First, we'll need to add an indicator variable for each sibling's frame 
population membership. In other words, we need to create a variable
that has the value 1 for each sibling who was eligible to respond to the 
survey, and 0 otherwise. In this example data, we'll assume that conditions 
similar to a typical DHS survey hold: i.e., we'll assume the survey design 
was such that siblings would have been eligible to respond if they

* are alive
* are female
* are between the ages of 15 and 50

For a given survey, these criteria will differ. You'll have to find out what 
the criteria for inclusion in the frame population were in order to produce 
estimates from sibling histories.

```{r}
ex.sib <- ex.sib %>% 
  mutate(in.F = as.numeric((sib.alive==1) & (sib.age >= 15) & (sib.age <= 49) & (sib.sex == 'f')))
```

Let's look at the distribution of frame population membership

```{r}
with(ex.sib, table(in.F, useNA='ifany'))
```

In this example dataset, a small number of the `in.F` values is missing.
(In a real dataset, there could well be more.)
Since we need to be able to determine whether or not each sibling is on in the
frame population, we would drop siblings missing `in.F` values from the analysis.

```{r}
ex.sib <- ex.sib %>% filter(! is.na(in.F)) 
```

### Specifying cells


When we produce estimated maternal death rates from sibling histories, we usually do so
for different sex X age group X time period combinations. These are called *cells*.

Our next step is to create an object that describes the cells that we plan to produce
death rate estimates for.
This means we need to specify the time period and age groups that we'll be using.
We'll use the helper function `cell_config` to do this:

```{r}
cc_mm <- cell_config(age.groups='5yr', 
                     time.periods='7yr_beforeinterview',
                     start.obs='sib.dob',    # date of birth
                     end.obs='sib.endobs',   # either the date respondent was interviewed (if sib is alive) or date of death (if sib is dead)
                     event='sib.maternal.death.date', # date of death (for sibs who died of maternal causes)
                     age.offset='sib.dob',   # date of birth
                     time.offset='doi',      # date of interview
                     exp.scale=1/12)
```

This step is where we process for estimating maternal death rates starts to 
differ from estimating all-cause death rates. 

We're specifying that we want

* 5-year age groups, starting from 15 and ending at 65
* to produce estimates for the 7-year window before each survey interview (so, slightly different for each respondent)
* the `dob` column of our sibling dataset has the time siblings start being observed (their birthdate)
* the `endobs` column of our sibling dataset has the time siblings stop being observed (the interview or, if they're dead, the date of death)
* the `sib.maternal.death.date` column of our sibling dataset has the date the sibling died of maternal causes (if the sibling died of maternal causes)
* the `doi` column of our sibling dataset has the date the respondent who reported about the sibling was interviewed
* in our survey, times are counted in months, so we set `exp.scale=1/12` to indicate that we need to divide total exposures by 12 to get years

### Estimating death rates

Given these preparatory steps, the `sibling_estimator` function will take care of estimating
death rates from the sibling histories for us.

```{r}
ex_ests <- sibling_estimator(sib.dat = ex.sib,
                             ego.id = 'caseid',            # column with the respondent id
                             sib.id = 'sibid',             # column with sibling id 
                                                           # (unique for each reported sibling)
                             sib.frame.indicator = 'in.F', # indicator for sibling frame population membership
                             sib.sex = 'sib.sex',          # column with sibling's sex
                             cell.config=cc_m,             # cell configuration we created above
                             weights='wwgt')               # column with the respondents' sampling weights

names(ex_ests)
```

`sibling_estimator` returns a list with the results. We'll focus on `asdr.ind`, which has the individual visibility
estimates, and `asdr.agg`, which has the aggregate visibility estimates.

Here are the individual visibility estimates:

```{r}
glimpse(ex_ests$asdr.ind)
```

And here are the aggregate visibility estimates

```{r}
glimpse(ex_ests$asdr.agg)
```

### Plotting the results

We'll make some plots showing the results


```{r}
ggplot(ex_ests$asdr.ind) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, group=sib.sex)) +
  theme_minimal() +
  scale_y_log10() + 
  ggtitle('individual visibility estimator, 7yr before survey')
```


```{r}
ggplot(ex_ests$asdr.agg) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, group=sib.sex)) +
  theme_minimal() +
  scale_y_log10() + 
  ggtitle('aggregate visibility estimator, 7yr before survey')
```

```{r}
compare <- bind_rows(ex_ests$asdr.ind, ex_ests$asdr.agg)

ggplot(compare) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, linetype=estimator, group=interaction(estimator, sib.sex))) +
  theme_minimal() +
  #facet_grid(sex ~ .) +
  facet_grid(. ~ sib.sex) +
  scale_y_log10() + 
  ggtitle('both estimators, 7yr before survey')

```

## Variance estimates

In practice, we want point estimates and estimated sampling uncertainty for the death rates.
We'll use the rescaled bootstrap to estimate sampling uncertainty. This can be done with the `surveybootstrap` package.

Before we use the rescaled bootstrap, we need to know a little bit about the sampling design of the survey
we're working with.
In this example dataset, we have a stratified, multistage design. So we'll need to tell the bootstrap function about
the strata and the primary sampling units. In this example data, these are indicated by the 'stratum' and 'psu'
columns of the dataset.

(NOTE: this takes about a minute or so on a 2018 MBP for 1000 resamples.)

```{r}
set.seed(101010)

tic('running bootstrap')
## this will take a little while -- for 1000 reps, it takes about 10 minutes on a 2018 Macbook Pro
#num.boot.reps <- 1000
# reduce number of reps to help vignette build faster
num.boot.reps <- 100


# The Guide to DHS Statistics DHS-8
# suggests using what we call the `stratum_design` variable
# (v023) in calculating sampling-based standard errors
bootweights <- surveybootstrap::rescaled.bootstrap.weights(survey.design = ~ psu + strata(stratum_design),
                                                           # a high number is good here, though that will obviously
                                                           # make everything take longer
                                                           num.reps=num.boot.reps,
                                                           idvar='caseid',       # column with the respondent ids
                                                           weights='wwgt', # column with the sampling weight
                                                           survey.data=ex.ego    # note that we pass in the respondent data, NOT the sibling data
                                                           )
toc()
```

The result, `bootweights`, is a dataframe that has a row for each survey respondent, a column for the survey respondent's ID,
and then `num.boot.reps` columns containing survey weights that result from the bootstrap procedure.
The basic idea is to calculate estimated death rates using each one of the `num.boot.reps` sets of weights.
The variation across the estimates is then an estimator for the sampling variation.

To make this easier, the `sibling_estimator` function can take a dataset with bootstrap resampled weights; it will
then calculate and summarize the estimates for you.

(NOTE: this is slow; it takes about 35 minutes or so on a 2018 MBP when `bootweights` has 1000 resamples.)

```{r}
# to save time, we'll only use a subset of the bootstrap replicates

short.bootweights <- bootweights %>% select(1:11)

#est.bootweights <- bootweights
est.bootweights <- short.bootweights

tic('calculating estimates with bootstrap')
ex_boot_ests <- sibling_estimator(sib.dat = ex.sib,
                                  ego.id = 'caseid',
                                  sib.id = 'sibid',
                                  sib.frame.indicator = 'in.F',
                                  sib.sex = 'sib.sex',
                                  cell.config=cc,
                                  boot.weights=est.bootweights,  # to get sampling uncertainty, we pass boot.weights into sibling_estimator
                                  return.boot=TRUE,                # when TRUE, return all of the resampled estimates (not just summaries)
                                  weights='wwgt')
toc()
```

Finally, let's plot the estimated death rates along with their sampling uncertainty:

```{r}
ggplot(ex_boot_ests$asdr.ind) +
  geom_ribbon(aes(x=sib.age, ymin=1000*asdr.hat.ci.low, ymax=1000*asdr.hat.ci.high, fill=sib.sex, group=sib.sex), alpha=.2) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, group=sib.sex)) +
  theme_minimal() +
  scale_y_log10() + 
  ggtitle('individual visibility estimator, 7yr before survey')
```

```{r}
ggplot(ex_boot_ests$asdr.agg) +
  geom_ribbon(aes(x=sib.age, ymin=1000*asdr.hat.ci.low, ymax=1000*asdr.hat.ci.high, fill=sib.sex, group=sib.sex), alpha=.2) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, group=sib.sex)) +
  theme_minimal() +
  scale_y_log10() + 
  ggtitle('aggregate visibility estimator, 7yr before survey')
```

## Internal consistency checks

In this section, we illustrate how to conduct internal consistency checks:

NOTE: The IC checks can take a while -- about 36 minutes for 1000 bootstrap reps on a 2018 MBP

```{r}
# toggle between short bootweights (faster, for coding) and long bootweights (for realism)
ic.bootweights <- short.bootweights
#ic.bootweights <- bootweights

tic("Internal consistency checks")
ic.checks <- sib_ic_checks(ex_boot_ests$esc.dat,
                           ego.dat=ex.ego,
                           ego.id='caseid',
                           sib.id='sibid',
                           sib.frame.indicator='in.F',
                           sib.cell.vars=c('sib.age', 'sib.sex'),
                           ego.cell.vars=c('age.cat', 'sex'),
                           boot.weights=ic.bootweights)
toc()

names(ic.checks)
```

We can look at `ic.checks$ic.summ`, which has summarized output:

```{r}
glimpse(ic.checks$ic.summ)
```



It's often helpful to plot the results of the internal consistency checks:

```{r}
ggplot(ic.checks$ic.summ) +
  geom_hline(yintercept=0) +
  geom_pointrange(aes(x=age.cat,
                      y=diff_mean,
                      ymin=diff_ci_low,
                      ymax=diff_ci_high)) +
  theme_minimal() +
  ggtitle("Internal consistency checks") +
  xlab("Age group") +
  ylab(expression(Delta[alpha]))

#ggplot(ic.checks$ic.summ) +
#  geom_hline(yintercept=0) +
#  geom_pointrange(aes(x=age.cat,
#                      y=agg_diff_mean,
#                      ymin=agg_diff_ci_low,
#                      ymax=agg_diff_ci_high)) +
#  theme_minimal() +
#  ggtitle("Internal consistency checks (aggregate)") +
#  xlab("Age group") +
#  ylab(expression(Delta[alpha]))
#
#ggplot(ic.checks$ic.summ) +
#  geom_hline(yintercept=0) +
#  geom_pointrange(aes(x=age.cat,
#                      y=ind_diff_mean,
#                      ymin=ind_diff_ci_low,
#                      ymax=ind_diff_ci_high)) +
#  theme_minimal() +
#  ggtitle("Internal consistency checks (individual)") +
#  xlab("Age group") +
#  ylab(expression(Delta[alpha]))
```

## Visibilities

```{r}
sib.F.dat <- ex.sib %>%
  group_by(caseid) %>%
  summarize(y.F = sum(in.F))

ego.vis <- ex.ego %>% 
  select(caseid, wwgt, age.cat, sex) %>%
  left_join(sib.F.dat, by='caseid')

# if nothing was joined in, there are no sibs
ego.vis <- ego.vis %>%
  mutate(y.F=ifelse(is.na(y.F), 0, y.F))

ego.vis.agg <- ego.vis %>%
  group_by(sex, age.cat) %>%
  summarise(y.F.bar = weighted.mean(y.F, wwgt)) %>%
  mutate(adj.factor = y.F.bar / (y.F.bar + 1))
```

Make adjusted individual estimates

```{r}
adj.agg.ests <- ex_ests$asdr.agg %>%
  left_join(ego.vis.agg,
            by=c('sib.sex'='sex', 'sib.age'='age.cat')) %>%
  mutate(adj.factor = ifelse(is.na(adj.factor), 1, adj.factor),
         estimator = 'sib_agg_adj',
         asdr.hat = asdr.hat * adj.factor)
```

And plot a comparison

```{r}
compare <- bind_rows(ex_ests$asdr.ind, ex_ests$asdr.agg, adj.agg.ests)

ggplot(compare) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, linetype=estimator, group=interaction(estimator, sib.sex))) +
  theme_minimal() +
  #facet_grid(sex ~ .) +
  facet_grid(. ~ sib.sex) +
  scale_y_log10() + 
  ggtitle('sibling estimators, 7yr before survey')

```


<!--
## TODO

* add calculation of visibilities by respondent, calculate avg vis by cell
* make it easier to produce adjusted estimates; demonstrate with S/(S+1) issue
* add normalization to IC checks
* make it easier to calculate IC checks by single year of age?
* right now, sibling_estimator now has hard-coded 'agelabel' references -- but it shouldn't? need to handle this more elegantly


# PROFILING

## TO PROFILE: BOOTSTRAP WEIGHTS

(NOTE: this takes about a minute or so on a 2018 MBP for 1000 resamples.)

```{r, eval=FALSE}
set.seed(101010)

tic('running bootstrap')
## this will take a little while -- for 1000 reps, it takes about 1 minutes on a 2018 Macbook Pro
#num.boot.reps <- 1000
# use a small number of reps to avoid the vignette taking forever to build
num.boot.reps <- 100
test.bootweights <- surveybootstrap::rescaled.bootstrap.weights(survey.design = ~ psu + stratum(stratum_analysis),
                                                           # a high number is good here, though that will obviously
                                                           # make everything take longer
                                                           num.reps=num.boot.reps,
                                                           idvar='caseid',       # column with the respondent ids
                                                           weights='wwgt', # column with the sampling weight
                                                           survey.data=ex.ego    # note that we pass in the respondent data, NOT the sibling data
                                                           )
toc()
```

-->
