---
title: "Working with MICS sibling history data"
author: "Dennis M. Feehan with collaboration of Claude (Opus 5)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with MICS sibling history data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

UNICEF's Multiple Indicator Cluster Surveys (MICS) collect sibling survival
histories in some rounds, using a module closely related to the one in the DHS.
`prep_mics_sib_histories()` prepares those data for the estimators in this
package.

This vignette covers what MICS data looks like, how it differs from the DHS, and
what we learned validating the package against published MICS estimates.

## Which MICS surveys can be used

MICS has fielded **two structurally different** mortality modules under the same
"MM" label, and only one of them is a sibling survival history.

| | MICS2 / MICS3 | MICS4 / MICS5 | MICS6 / MICS7 |
|---|---|---|---|
| Method | Indirect sisterhood | **Direct sibling history** | **Direct sibling history** |
| Usable here? | No | Yes | Yes |
| Respondents | All adults 15+ | Women 15--49 | Women 15--49 |
| Brothers enumerated? | No | Yes | Yes |
| Roster numbering | --- | `MM5`--`MM13` | `MM15`--`MM27` |
| Estimand | --- | Pregnancy-related | Maternal *and* pregnancy-related |

MICS2 and MICS3 collect summary sisterhood counts in the *household*
questionnaire -- no per-sibling roster, no ages at death -- so they cannot be
used with this package at all.

The direct sibling module is optional, and most surveys did not field it. The practical
test is whether the survey ships an `mm.sav` file. Of 229 MICS surveys we
examined, only 13 have a usable sibling history: 2 in MICS4, 7 in MICS5 and 4 in
MICS6.

Two traps when building your own inventory:

* **MICS3's `mm.sav` is the *men's* file**, not maternal mortality. It contains
  variables like `MM4` "Man's Line number" and `mmweight` "men's sample weight".
  Globbing for `mm.sav` will pick these up.
* Surveys that did not field the module do **not** hide the roster in `wm.sav`
  -- there simply are no `MM` variables.

You can check a survey before downloading it: the World Bank microdata catalog
publishes full variable dictionaries without registration.

## A minimal example

MICS publishes the roster as one row per reported sibling, so no reshaping is
needed.

```{r, eval=FALSE, purl=FALSE}
library(siblingsurvival)
library(haven)

mm <- read_sav("mm.sav")

prepped <- prep_mics_sib_histories(mm, survey = "IQ2018")

prepped$ego.dat   # one row per respondent
prepped$sib.dat   # one row per reported sibling
prepped$summ      # a one-row summary
```

`survey` should have a code that identifies the survey -- e.g., `ZW2014`. 
DHS surveys include this code in the datafiles, but MICS surveys do not,
so you have to supply it here.

For MICS4 or MICS5 data, pass the matching varmap:

```{r, eval=FALSE, purl=FALSE}
prep_mics_sib_histories(mm, survey = "ZW2014", varmap = sibhist_varmap_mics5)
```

Some countries omit the respondent's date of birth from `mm.sav`. When that
happens, supply the women's file so respondent age can be recovered:

```{r, eval=FALSE, purl=FALSE}
prep_mics_sib_histories(mm, survey = "MDG2018", wm.df = read_sav("wm.sav"))
```

## Special features of MICS files 

MICS files differ from DHS files in several ways; `prep_mics_sib_histories()` handles 
all of the following.

**Weights are already normalized.** The DHS publishes women's weights multiplied
by 1,000,000; MICS `wmweight` column, on the other hand, averages 1. The MICS
prep therefore defaults to `weight.scale = 1`. Passing DHS-style scaling here
would divide every weight by a million, and nothing downstream would complain.

**Survival status is coded differently.** MICS codes `MM16` as 1 yes / 2 no /
8 don't know, while this package follows the DHS convention of 1 alive / 0 dead.
The prep file recodes this.

**Don't-know values are codes, not blanks.** MICS uses **98** and **99** on the
numeric items (`MM17` age, `MM18` years since death, `MM19` age at death). Left
as-is, a sibling with `sib.death.age = 98` gets a date of birth 98 years before
her death. This is masked whenever MICS supplies its own imputed CMC dates, so it
only affects surveys that ship no imputed CMC dates. The prep function will set 
these to `NA`.

**`MM16` collides with the DHS.** In MICS6, the `mm16` variable is the survey question
that asks "Is (name) still alive?"; in DHS phase 7 and later `mm16` is the survey
question that asks "died of violence or an accident". To avoid errors resulting from
confusing these two, the prep function refuses to run a varmap that maps `mm16` to 
`sib.died.accident`.

**Several columns must be constructed.** MICS has no single respondent id, so
`caseid` is built from cluster, household and line number. `doi` comes from
`WDOI` where present and is otherwise built from the interview year and month;
it must end up a century month code, since all the date derivations are integer
arithmetic in months. Respondent `age` is derived from `WDOI` and `WDOB`, `psu`
falls back to the cluster (only 4 of the 13 usable surveys carry a `psu` column),
and `sex` is set to `'f'`, since MICS interviews only women.

**Names are mixed case.** Questionnaire items and link keys are uppercase
(`MM15`, `HH1`, `WDOI`) while derived and design variables are lowercase
(`wmweight`, `psu`, `welevel`). The varmaps are lowercase, so
`lowercase = TRUE` is used to normalize the input.

**Sibling dates may or may not be supplied.** MICS6 usually includes `MM17C` and
`MM18C` (imputed CMC birth and death dates), and MICS4/5 includes `MM7C` and `MM8C`,
but some countries include neither. The package uses them when present and
approximates them from reported ages and years-since-death otherwise. 

## Validation against published estimates

We evaluated code in the package by comparing its estimates to published tables
from three MICS6 surveys and one MICS5 survey.

It is worth separating two kinds of quantity, because they test different
things.

### Counts: are we reading the data correctly?

Exposure and death counts are essentially sums over survey responses. Getting
them right means the varmap, the recodes, the reference window, the weights and
the age placement are all correct -- and nothing else. Ratios of our figure to
the published one:

| | Iraq 2018 | Madagascar 2018 | Zimbabwe 2014 | Zimbabwe 2019 |
|---|---|---|---|---|
| Round | MICS6 | MICS6 | **MICS5** | MICS6 |
| Female exposure | 1.0000 | 1.0003 | 0.9995 | 0.997 |
| Male exposure | 0.9993 | 1.0003 | 0.9994 | 0.999 |
| Female all-cause deaths | 1.0007 | 0.9994 | 1.0037 | 0.987 |
| Male all-cause deaths | 0.9996 | 1.0003 | 1.0015 | 1.002 |
| Pregnancy-related deaths | 1.008 | 1.025 | 0.982 | 0.923 |

### Rates: does the estimator agree?

These add the parts the counts do not exercise -- age standardisation by the
respondents' age distribution, and the life-table conversion behind
$_{35}q_{15}$. Our figure, with the published one in brackets:

| | Iraq 2018 | Madagascar 2018 | Zimbabwe 2014 | Zimbabwe 2019 |
|---|---|---|---|---|
| All-cause rate, women$^a$ | 1.25 (1.25) | 3.08 (3.1) | 9.98 (9.84) | 6.27 (6.28) |
| All-cause rate, men$^a$ | 2.39 (2.39) | 4.03 (4.0) | 9.90 (9.79) | 6.16 (6.09) |
| Pregnancy-related rate$^a$ | 0.135 (0.13) | 0.67 (0.65) | 0.84 (0.86) | 0.54 (0.59) |
| $_{35}q_{15}$, women$^b$ | 49.5 (49) | 111.9 (112) | 345 (344) | 222.7 (224) |
| $_{35}q_{15}$, men$^b$ | 85.8 (86) | 148.8 (149) | 357 (357) | 218.5 (219) |

$^a$ per 1,000 person-years, age-adjusted. $^b$ per 1,000.

Where the counts agree the rates agree too, which is the expected pattern: the
standardisation and life-table steps add no error of their own.

Most of the estimates reproduce published values very closely, across both roster schemes and
three countries. The discrepancy in pregnancy related deaths with Zimbabwe 2019 
are larger than for the other three surveys;
we have been unable to determine exactly why this is the case.

<!--
it misses on quantities that involve no maternal recode at all, so the problem is
in that survey's published tables rather than in the estimator. Zimbabwe's report
is independently known to be internally inconsistent: its sampling-error appendix
gives a maternal mortality ratio of 413.6 against its own headline figure of 462,
whereas Iraq's sampling-error table agrees with its main table. Note that
Zimbabwe *2014* reproduces exactly, so this is specific to the 2019 survey's
published tables rather than to the country.
-->

**The target is the aggregate-visibility estimator.** `asdr.agg` is
`sum(w * deaths) / sum(w * exposure)` with no visibility weighting, which is
exactly the conventional MICS calculation -- MICS applies no sibship-size
correction. `asdr.ind`, the individual-visibility estimator, is *expected* to
differ from published values.


All 13 MICS surveys with a sibling roster have been run through the prep. Every
one completes, produces no `NA` estimates, and gives a plausible
pregnancy-related rate (0.08 to 1.39 per 1,000 woman-years). Two turned up
country-specific data quirks worth knowing about:

* **São Tomé and Príncipe 2014** has one respondent with no sampling weight, and
  names its weight column `WMWEIGHT` rather than `wmweight`.
* **Bhutan 2010** contains two siblings whose reported years-since-death and age
  at death are jointly inconsistent -- one died 58 years ago at age 58, implying
  a birth 116 years before the interview. The prep warns rather than silently
  carrying it. Note this is *not* a check on whether a sibling died before the
  respondent was born, which is perfectly possible and not rare where fertility
  is high and sibships are long.

Also, two conventions that MICS documents do not state were settled by this exercise.

**The reference period is the seven years *including* the year of interview.**
Testing candidate windows against Iraq's and Zimbabwe's published exposure,
`[doi - 84, doi)` matches while `[doi - 96, doi - 12)` does not. This is what
`cell_config(time.periods = '7yr_beforeinterview')` already does.

**Age standardisation uses the age distribution of the interviewed women**, which
is what `get_ego_age_distn()` computes -- not household women, as some secondary
sources state. Standardising Zimbabwe's all-cause female rate this way gives 6.27
against a published 6.28.

### Table TM.9.3 counts pregnancy-related deaths

This one matters for interpretation.

MICS survey reports print a table headed *Maternal mortality*, with a column
labelled **"Maternal Deaths"** and a footnote defining a maternal death as one
"from any cause except accidents or violence". **The number printed is not that
quantity.** It is the *pregnancy-related* count: deaths during pregnancy,
childbirth, or within two months of the end of a pregnancy, with **no** cause
exclusion and **no** 42-day cut.

The reports' own methodology text says so -- "age-specific mortality rates are
calculated by dividing the number of **pregnancy-related deaths** by years of
exposure" -- and the arithmetic confirms it. In Iraq the pregnancy-related count
reproduces the published figure at ratio 1.008 while the maternal count gives
0.848; in Madagascar, 1.025 against 0.792.

So when comparing against published MICS figures, use
`sib.preg_related.death.date`, not `sib.maternal.death.date`, despite what the
published column is called.

### What will not reproduce, by design

* **`asdr.ind`.** MICS applies no sibship-size correction, so the
  individual-visibility estimator differs from published figures by construction.
* **Confidence intervals.** MICS uses jackknife replication with $\pm 2 \cdot se$;
  this package uses a bootstrap with percentile intervals.
* **The maternal mortality ratio itself.** MMR is the maternal mortality rate
  divided by a general fertility rate computed from the women's birth history --
  a different module this package does not read.
Rows that cannot contribute to a weighted estimate -- no usable date of birth,
or no sampling weight -- are dropped and counted in `summ$miss.dob` and
`summ$miss.wgt`. This matters more than it sounds: a single such row turns an
entire estimate cell into `NA`, because the estimator sums over the cell.

* **Exposure, to the last person-year.** Siblings whose survival status is
  unknown are dropped by default, and MICS appears to censor rather than drop
  them: published exposure sits between our `keep_missing = FALSE` and
  `keep_missing = TRUE` values. The spread is under 0.1% and does not move any
  published rate to two decimal places.

## Maternal versus pregnancy-related

MICS6 and MICS7 ask whether the death was due to violence (`MM26`) or an accident
(`MM27`), so both estimands are computable. MICS4 and MICS5 do not, so they
support pregnancy-related mortality only -- the same limitation as DHS phases
2--6.

Because MICS6 also records the number of days after the end of pregnancy
(`MM25`), the older two-month pregnancy-related series can be reconstructed from
MICS6 microdata. That is what makes MICS6 comparable with DHS surveys and with
earlier MICS rounds.

Three questionnaire skip patterns must be respected when classifying deaths:

* `MM23 = 1` (died during childbirth) skips `MM26` and `MM27` entirely, so those
  are `NA` **by design**. A childbirth death is unconditionally maternal and must
  not be excluded for missing cause information.
* `MM21` routes sisters who died before age 12 past `MM22`--`MM25`, so those
  items are also `NA` by design. An explicit age guard is needed, or every
  under-12 female death is counted as maternal.
* `MM16 = DK` jumps to the next sibling with no age or date information at all.
