---
title: "Working with MICS sibling history data"
author: "Dennis M. Feehan with collaboration of Claude (Opus 5)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with MICS sibling history data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

UNICEF's Multiple Indicator Cluster Surveys (MICS) collect sibling survival
histories in some rounds, using a module closely related to the one in the DHS.
`prep_mics_sib_histories()` prepares those data for the estimators in this
package.

This vignette covers what MICS data looks like, how it differs from the DHS, and
what we learned validating the package against published MICS estimates.

## Which MICS surveys can be used

MICS has fielded **two structurally different** mortality modules under the same
"MM" label, and only one of them is a sibling survival history.

| | MICS2 / MICS3 | MICS4 / MICS5 | MICS6 / MICS7 |
|---|---|---|---|
| Method | Indirect sisterhood | **Direct sibling history** | **Direct sibling history** |
| Usable here? | No | Yes | Yes |
| Respondents | All adults 15+ | Women 15--49 | Women 15--49 |
| Brothers enumerated? | No | Yes | Yes |
| Roster numbering | --- | `MM5`--`MM13` | `MM15`--`MM27` |
| Estimand | --- | Pregnancy-related | Maternal *and* pregnancy-related |

MICS2 and MICS3 collect summary sisterhood counts in the *household*
questionnaire -- no per-sibling roster, no ages at death -- so they cannot be
used with this package at all.

The direct sibling module is optional, and most surveys did not field it. The practical
test is whether the survey ships an `mm.sav` file. Of 229 MICS surveys we
examined, only 13 have a usable sibling history: 2 in MICS4, 7 in MICS5 and 4 in
MICS6.

Two traps when building your own inventory:

* **MICS3's `mm.sav` is the *men's* file**, not maternal mortality. It contains
  variables like `MM4` "Man's Line number" and `mmweight` "men's sample weight".
  Globbing for `mm.sav` will pick these up.
* Surveys that did not field the module do **not** hide the roster in `wm.sav`
  -- there simply are no `MM` variables.

You can check a survey before downloading it: the World Bank microdata catalog
publishes full variable dictionaries without registration.

## A minimal example

MICS publishes the roster as one row per reported sibling, so no reshaping is
needed.

```{r, eval=FALSE, purl=FALSE}
library(siblingsurvival)
library(haven)

mm <- read_sav("mm.sav")

prepped <- prep_mics_sib_histories(mm, survey = "IQ2018")

prepped$ego.dat   # one row per respondent
prepped$sib.dat   # one row per reported sibling
prepped$summ      # a one-row summary
```

`survey` should have a code that identifies the survey -- e.g., `ZW2014`. 
DHS surveys include this code in the datafiles, but MICS surveys do not,
so you have to supply it here.

For MICS4 or MICS5 data, pass the matching varmap:

```{r, eval=FALSE, purl=FALSE}
prep_mics_sib_histories(mm, survey = "ZW2014", varmap = sibhist_varmap_mics5)
```

Some countries omit the respondent's date of birth from `mm.sav`. When that
happens, supply the women's file so respondent age can be recovered:

```{r, eval=FALSE, purl=FALSE}
prep_mics_sib_histories(mm, survey = "MDG2018", wm.df = read_sav("wm.sav"))
```

## Special features of MICS files 

MICS files differ from DHS files in several ways; `prep_mics_sib_histories()` handles 
all of the following.

**Weights are already normalized.** The DHS publishes women's weights multiplied
by 1,000,000; MICS `wmweight` column, on the other hand, averages 1. The MICS
prep therefore defaults to `weight.scale = 1`. Passing DHS-style scaling here
would divide every weight by a million, and nothing downstream would complain.

**Survival status is coded differently.** MICS codes `MM16` as 1 yes / 2 no /
8 don't know, while this package follows the DHS convention of 1 alive / 0 dead.
The prep file recodes this.

**Don't-know values are codes, not blanks.** MICS uses **98** and **99** on the
numeric items (`MM17` age, `MM18` years since death, `MM19` age at death). Left
as-is, a sibling with `sib.death.age = 98` gets a date of birth 98 years before
her death. This is masked whenever MICS supplies its own imputed CMC dates, so it
only affects surveys that ship no imputed CMC dates. The prep function will set 
these to `NA`.

**`MM16` collides with the DHS.** In MICS6, the `mm16` variable is the survey question
that asks "Is (name) still alive?"; in DHS phase 7 and later `mm16` is the survey
question that asks "died of violence or an accident". To avoid errors resulting from
confusing these two, the prep function refuses to run a varmap that maps `mm16` to 
`sib.died.accident`.

**Several columns must be constructed.** MICS has no single respondent id, so
`caseid` is built from cluster, household and line number. `doi` comes from
`WDOI` where present and is otherwise built from the interview year and month;
it must end up a century month code, since all the date derivations are integer
arithmetic in months. Respondent `age` is derived from `WDOI` and `WDOB`, `psu`
falls back to the cluster (only 4 of the 13 usable surveys carry a `psu` column),
and `sex` is set to `'f'`, since MICS interviews only women.

**Names are mixed case.** Questionnaire items and link keys are uppercase
(`MM15`, `HH1`, `WDOI`) while derived and design variables are lowercase
(`wmweight`, `psu`, `welevel`). The varmaps are lowercase, so
`lowercase = TRUE` is used to normalize the input.

**Sibling dates may or may not be supplied.** MICS6 usually includes `MM17C` and
`MM18C` (imputed CMC birth and death dates), and MICS4/5 includes `MM7C` and `MM8C`,
but some countries include neither. The package uses them when present and
approximates them from reported ages and years-since-death otherwise. 

## Validation against published estimates

We evaluated code in the package by comparing its estimates to published tables
from three MICS6 surveys and one MICS5 survey.

First, we examing quantities that are essentially sums over survey responses.
The following table shows ratios of the value produced by the package to the 
value published in the MICS report:

| | Iraq 2018 | Madagascar 2018 | Zimbabwe 2014 | Zimbabwe 2019 |
|---|---|---|---|---|
| Round | MICS6 | MICS6 | **MICS5** | MICS6 |
| Female exposure | 0.9999 | 1.0003 | 0.9995 | 0.9966 |
| Male exposure | 0.9993 | 1.0003 | 0.9994 | 0.9990 |
| Female all-cause deaths | 1.0007 | 0.9994 | 1.0037 | 0.9877 |
| Male all-cause deaths | 0.9996 | 1.0003 | 1.0015 | 1.0023 |
| Pregnancy-related deaths | 1.0068 | 0.9967 | 1.0002 | 0.8962 |

The pregnancy-related row uses the postpartum window each round's own published
tables use: `preg.window = "42days"` for the three MICS6 surveys and the
two-month default for Zimbabwe 2014, since MICS5 asks no day count and its
reports use the wider definition. The next section explains why.

The values are generally very close, with the exception of the count of
pregnancy-related deaths in the Zimbabwe 2019 survey, which is discussed below.

Next, we examined the extent to which estimated mortality rates agree between
the package and the MICS reports.
These quantities are more complex; they are ratios, some involve standarizing by the respondents'
age distribution, and some involve life table calculations. 
Note that these rates all come from the aggregate visibility estimator; the
individual visibility estimator has not been used in any MICS reports.

This table shows the estimates produced by the package, along with estimates
published in MICS reports in parentheses:

| | Iraq 2018 | Madagascar 2018 | Zimbabwe 2014 | Zimbabwe 2019 |
|---|---|---|---|---|
| All-cause rate, women$^a$ | 1.25 (1.25) | 3.08 (3.1) | 9.98 (9.84) | 6.28 (6.28) |
| All-cause rate, men$^a$ | 2.39 (2.39) | 4.03 (4.0) | 9.90 (9.79) | 6.16 (6.09) |
| Pregnancy-related rate$^a$ | 0.133 (0.13) | 0.650 (0.65) | 0.862 (0.86) | 0.530 (0.59) |
| $_{35}q_{15}$, women$^b$ | 49.5 (49) | 111.9 (112) | 345.0 (344) | 222.8 (224) |
| $_{35}q_{15}$, men$^b$ | 85.8 (86) | 148.8 (149) | 356.7 (357) | 218.4 (219) |

Here, estimates produced by the package match estimates in the MICS reports quite
closely.

## What the official tabulation syntax says

UNICEF distributes the SPSS syntax used to produce the MICS6 report tables. One
file, `MICS6 - 06 - TM.9.1&TM.9.2&TM.9.3&DQ.7.1&DQ7.2.sps`, produces the whole
mortality group. Transcribing it literally into R -- see
`data-raw/mics-validation/spss-syntax-replica.R` -- settles several questions
that the reports themselves leave open, and it reproduces Iraq 2018 and
Madagascar 2018 **to the person-year**:

| | Iraq 2018 | Madagascar 2018 |
|---|---|---|
| Female exposure | 471,294 (471,294) | 202,959 (202,959) |
| Male exposure | 485,398 (485,398) | 208,274 (208,274) |
| Female all-cause deaths | 574.4 (574) | 616.6 (617) |
| Pregnancy-related deaths | 64.4 (64) | 136.6 (137) |
| General fertility rate | 128.2 (128) | --- |
| Maternal mortality ratio | 104.0 (104) | --- |

### The published "maternal deaths" column is a 42-day pregnancy-related count

The syntax flags a death with

```
if (MM15 = 2 & (MM22 = 1 or MM23 = 1 or (MM24 = 1 and MM25 < 42))) md = 1.
```

`MM26` (violence) and `MM27` (accident) are **never read anywhere in the file**,
even though the reports carry a footnote saying deaths from those causes are
excluded. So what is printed under "Maternal Deaths" is a *pregnancy-related*
count restricted to a 42-day postpartum window -- which is the WHO definition of
a pregnancy-related death, and is neither of the two things this package used to
compute. On Iraq 2018 the three candidate definitions give:

| Definition | Deaths | Ratio to published 64 |
|---|---|---|
| `MM22 \| MM23 \| MM24` -- two months, any cause | 67.7 | 1.057 |
| `MM22 \| MM23 \| (MM24 & MM25 < 42)` -- 42 days, any cause | **64.4** | **1.007** |
| ... additionally excluding `MM26`/`MM27` -- our maternal column | 59.0 | 0.922 |

`add_maternal_deaths(preg.window = "42days")` selects the middle row, and is what
you want when comparing against a published MICS table. The default stays
`"2months"`, which is the wider DHS-comparable reading. Note that the 42-day cut
is also what the package has always applied on the DHS side, through the `mm12`
band `100`--`141`.

### Other conventions the syntax fixes

* The seven-year window is `[wdoi - 84, wdoi - 1]`: the interview month is
  excluded. `cell_config(time.periods = '7yr_beforeinterview')` already does this.
* Age standardisation uses the **weighted** age distribution of women with a
  completed interview (`WM17 = 1`), which is what `get_ego_age_distn()` computes.
* A decedent's exposure ends the month *before* death (`MM18C - 1`), and the
  death is assigned to the age group holding at that month rather than to the
  age at death.
* Siblings whose survival status is unknown are **kept**, contributing full
  exposure. This package drops them instead, which is why our exposure totals sit
  within about 0.1% of the published ones rather than exactly on them --- Iraq
  has 107 such siblings, Zimbabwe 2019 has 42, Madagascar only 9.
* $_{35}q_{15}$ is computed with `1 + 2.4*nmx` in the denominator, i.e. $n_ax =
  2.6$, although the same file's header documents $n_ax = 2.5$. The difference is
  at most 0.2 per 1,000 and cannot be resolved against published figures, which
  are rounded to integers.

### Zimbabwe 2019

Running the official syntax on Zimbabwe 2019's own public microdata reproduces
its **male** columns and fails on its **female** ones:

| | Ours | Published | Ratio |
|---|---|---|---|
| Male exposure | 110,072 | 110,089 | 0.9998 |
| Male all-cause deaths | 669.5 | 668 | 1.002 |
| Male all-cause rate | 6.087 | 6.09 | 0.9995 |
| Female exposure | 108,644 | 108,985 | 0.9969 |
| Female all-cause deaths | 665.7 | 674 | 0.988 |
| Pregnancy-related deaths | 60.9 | 68 | 0.896 |

Both sexes go through identical code, differing only in which value of `MM15`
they select on, so a data-handling difference that hits one and not the other is
not something the calculation can produce. No variation on the window, the age
restriction, the weighting, or the treatment of the 17 sex-missing rows brings
the female figures up to the published ones.

The report also contradicts itself. Its sampling-error appendix gives a maternal
mortality ratio (indicator TM.21) of **413.64**, against **462** in TM.9.3.
Those two figures come from different implementations -- the appendix notes that
maternal mortality ratio errors "can be calculated only through the separate CS
Pro application" -- and it is the appendix that we reproduce: the official syntax
on the public microdata gives 409.9, within 1% of 413.64 and 11% below the
headline. The fertility side of the same table reproduces exactly (our GFR 128.4
against a published 128), so the discrepancy is confined to the female sibling
records.

Zimbabwe *2014* reproduces essentially exactly -- its pregnancy-related death
count comes out at 1.0002 of the published figure -- so this is specific to the
2019 survey's published tables rather than to the country. Treat Zimbabwe 2019's published
female mortality figures as unreliable.

## Running every survey through the prep

All 13 MICS surveys with a sibling roster have been run through the prep. Every
one completes, produces no `NA` estimates, and gives a plausible
pregnancy-related rate (0.07 to 1.46 per 1,000 woman-years). Two turned up
country-specific data quirks worth knowing about:

* **São Tomé and Príncipe 2014** has one respondent with no sampling weight, and
  names its weight column `WMWEIGHT` rather than `wmweight`. That single
  weightless respondent used to make `sum(wwgt)` `NA` inside
  `get_ego_age_distn()`, which propagated through the denominator and turned
  *every* group proportion `NA` -- silently NA-ing out every age-standardised
  rate for the survey. Such respondents are now dropped, with a warning.
* **Bhutan 2010** contains two siblings whose reported years-since-death and age
  at death are jointly inconsistent -- one died 58 years ago at age 58, implying
  a birth 116 years before the interview. The prep function warns rather than silently
  carrying it. 

Rows that cannot contribute to a weighted estimate -- no usable date of birth, or
no sampling weight -- are dropped, and counted in `summ$miss.dob` and
`summ$miss.wgt`.

### What will not necessarily agree with reports, by design

* **`asdr.ind`.** This estimator is not used in MICS reports.
* **Siblings of unknown survival status.** The package drops them; the official
  syntax keeps them at full exposure. This is the main reason exposure totals
  land within about 0.1% of published ones rather than exactly on them.
* **Confidence intervals.** MICS uses jackknife replication with $\pm 2 \cdot se$;
  this package uses a bootstrap with percentile intervals.
* **The maternal mortality ratio itself.** MMR is the maternal mortality rate
  divided by a general fertility rate computed from the women's birth history --
  a different module this package does not read. (The replica script does compute
  it, for validation purposes only.)

## Maternal versus pregnancy-related

MICS6 and MICS7 ask whether the death was due to violence (`MM26`) or an accident
(`MM27`), so both estimands are computable. MICS4 and MICS5 do not, so they
support pregnancy-related mortality only -- the same limitation as DHS phases
2--6.

Because MICS6 also records the number of days after the end of pregnancy
(`MM25`), the older two-month pregnancy-related series can be reconstructed from
MICS6 microdata. That is what makes MICS6 comparable with DHS surveys and with
earlier MICS rounds.

There are really three quantities in play, and it is worth being explicit about
which one is wanted:

| | Postpartum window | Excludes violence/accident? | How to get it |
|---|---|---|---|
| Pregnancy-related, DHS-comparable | two months | no | `preg.window = "2months"` (default) |
| Pregnancy-related, WHO / published MICS | 42 days | no | `preg.window = "42days"` |
| Maternal | 42 days | yes | `sib.maternal.death.date` |

The middle row is what MICS6 and MICS7 reports print, however they are labelled.

MICS4 and MICS5 never ask `MM25`, so only the first row is available for them ---
and it is also the right one, since the syntax's own header notes that in earlier
rounds "maternal deaths were defined as deaths of women during pregnancy or
childbirth or within two months after the termination of a pregnancy or
childbirth, inclusive of all causes". Asking for `preg.window = "42days"` on
MICS4/5 data is an error rather than a silent no-op.

Three questionnaire skip patterns must be respected when classifying deaths:

* `MM23 = 1` (died during childbirth) skips `MM26` and `MM27` entirely, so those
  are `NA` **by design**. A childbirth death is unconditionally maternal and must
  not be excluded for missing cause information.
* `MM21` routes sisters who died before age 12 past `MM22`--`MM25`, so those
  items are also `NA` by design. An explicit age guard is needed, or every
  under-12 female death is counted as maternal.
* `MM16 = DK` jumps to the next sibling with no age or date information at all.
