---
title: "Preparing DHS data"
author: "Dennis M. Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



```{r}
library(siblingsurvival)
library(tidyverse)

# this is helfpul for timing
library(tictoc)
```

## Overview

This vignette shows how to prepare a DHS-style dataset for analysis with the
`siblingsurvival` model.

First, we'll open up the model DHS dataset, which is available
[here](https://dhsprogram.com/customcf/legacy/data/sample_download_dataset.cfm?Filename=ZZIR62DT.ZIP&Tp=1&Ctry_Code=zz&survey_id=0&doctype=dhs);
see also [this webpage](https://dhsprogram.com/data/Download-Model-Datasets.cfm).
We're going to be using the **Individual Recode**, and, generally, it's easiest to start
using the Stata-formatted version (neither an R version, nor an easy-to-use plain text format like .csv is provided).

We'll assume that we've read the individual recode version of the dataset into R using `haven::read_dta`,
and that we're calling it `model_dhs_dat`.

```{r}
data(model_dhs_dat)
```

Next, we'll open up the `varmap`, which maps column names in the raw DHS data to more
user-friendly names.

An up-to-date list of DHS versions and recodes is available [here](https://www.dhsprogram.com/publications/publication-dhsg4-dhs-questionnaires-and-manuals.cfm).

```{r}
data(sibhist_varmap_dhs6)
```

Using the `varmap`, get a vector with the names of the respondent (ego) variables

```{r}
## ego (respondent) variables to grab
tmp <- subset(sibhist_varmap_dhs6, sibvar==0)
resp.attrib <- tmp$orig.varname
names(resp.attrib) <- tmp$new.varname
resp.attrib
```

Using the `varmap`, get a vector with the names of the sibling variables

```{r}
## alter (sibling) variables to grab
tmp <- subset(sibhist_varmap_dhs6, sibvar==1)
sib.attrib <- tmp$orig.varname
names(sib.attrib) <- tmp$new.varname
sib.attrib
```

We'll need to prepare two different datasets: the first dataset has information about the women
who responded to the survey; we'll call this the respondent or ego dataset.
The second dataset has information about the reported siblings; we'll call this the sibling dataset.

## Prepare the respondent (ego) dataset

```{r}
ex.ego <- model_dhs_dat %>% 
  # use information from the varmap to rename ego variables
  rename(!!!resp.attrib) %>%
  mutate(
    ## typically, only women are asked sibling histories in DHS surveys
    sex='f',
    ## these breaks should agree with the ones
    ## used in the age categories for siblings
    ## (see below)
    age.cat=fct_drop(cut(age,
      breaks=c(0, seq(from=15,to=50,by=5),95),
      include.lowest=TRUE, right=FALSE)),
    ## we'll rescale the weights, dividing them by 1,000,000 
    ## (so that their average is 1); see DHS documentation
    wwgt=wwgt/1e6
  ) 

ex.ego
```

## Prepare the sibling reports dataset

Next, we'll prepare the dataset of sibling reports.

We'll start by using `attributes.to.long` to convert the wide-form sibling data into long format for analysis:

```{r}
sibdata <- siblingsurvival::attributes.to.long(ex.ego,
                                               attribute.prefix=sib.attrib,
                                               ego.vars=c('caseid', 'wwgt', 
                                                          'psu', 'doi'),
                                               idvar="caseid")
glimpse(sibdata)
```

Now we'll do some recoding.

Turn the `sib.sex` variable into more readable values:

```{r}
sibdata.coded <- sibdata %>%
  mutate(sib.sex = ifelse(sib.sex == 2, 'f', 'm'))
```

Next, we need to create a new variable, `sib.endobs`, which has the
date (as a century month code, CMC) when we stopped observing each sibling.
Generally, this date is either the time that the sibling died, or the date
of the interview.

```{r}
## make the assumption that
##  (1) sibs who died lived all the way through
##      the month in which they are reported to have died
##  (2) the interview took place on the first of the month
## these assumptions are necessary to get these tables to line up
## with the DHS reports
sibdata.coded$sib.endobs <- pmin(sibdata.coded$doi,
                                 sibdata.coded$sib.death.date + 1,
                                 na.rm=TRUE)
```

For siblings who are currently alive, we'll recode their date of death to -1;
this ensures that we count their exposure, but not their deaths (since they are alive).

```{r}
## siblings who haven't died get their death dates
## recoded to -1 so we don't lose the exposures they
## contribute...
sibdata.coded$sib.death.date[ is.na(sibdata.coded$sib.death.date) ] <- -1
```

We'll also create a unique id for each reported sibling.

```{r}
sibdata.coded$sibid <- 1:nrow(sibdata)
```

Now we'll do some quality checks, removing some sibling reports when there is not enough
information to make use of them.

First, we'll figure out how many siblings have unknown survival status, and we'll
drop these.

```{r}

## CALCULATE % of siblings with unknown survival status;
## we are taking them out of the analysis here
sibdata.coded %>% 
  group_by(sib.alive) %>% 
  summarise(freq=n()) %>% 
  mutate(frac=freq/sum(freq))

pre.n <- nrow(sibdata.coded)
sibdata.coded <- sibdata.coded %>% filter(sib.alive %in% c(0,1))
post.n <- nrow(sibdata.coded)

cat(paste0("Removed ", pre.n-post.n, " out of ", pre.n, " (", round(100*(pre.n-post.n)/pre.n,2), "%)",
           " reports about sibs with unknown survival status.\n"))
```

Next, we'll figure out how many siblings have unknown sex, and we'll drop them, too.

```{r}

## CALCULATE % of siblings with unknown sex
sibdata.coded %>% 
  group_by(sib.sex) %>% 
  summarise(freq=n()) %>% 
  mutate(frac=freq/sum(freq))

## we're removing them from the analysis here
pre.n <- nrow(sibdata.coded)
sibdata.coded <- sibdata.coded %>% filter(sib.sex %in% c('f', 'm'))
post.n <- nrow(sibdata.coded)

cat(paste0("Removed ", pre.n-post.n, " out of ", pre.n, " (", round(100*(pre.n-post.n)/pre.n,2), "%)",
           " reports about sibs with unknown sex.\n"))

ex.sib <- sibdata.coded
```

We'll only keep a small subset of the ego variables, to keep things manageable

```{r}
ex.ego <- ex.ego %>% select(!!!c(names(resp.attrib), 'sex', 'age.cat'))
```

## Done!

Now our two datasets, `ex.ego` and `ex.sib` are prepped and ready to go!

# Using `prep_dhs_sib_histories`

The package includes a function, `prep_dhs_sib_histories`, that can be used
to load the sibling data, following the steps we just described. It works like this:

```{r}
data(model_dhs_dat)
data(sibhist_varmap_dhs6)

prepped <- prep_dhs_sib_histories(model_dhs_dat,
                                  varmap = sibhist_varmap_dhs6,
                                  keep_missing = FALSE)

str(prepped)
```


