---
title: "Estimating death rates from sibling history data"
author: "Dennis M. Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(siblingsurvival)
library(tidyverse)
```


We'll open up the demo datsets. First, here's a dataset that has information about survey respondents. (We'll refer to these survey respondents as 'ego':

```{r}
data(ex.ego)
ex.ego
```

And here's a long-form version of the sibling history data -- there's one row for each reported sibling

```{r}
data(ex.sib)
ex.sib
```

## Overview

It will be helpful to define a few important terms before we start our analysis:

* `ego` - an ego is a survey respondent
* `cell` - a cell is a generic group for which we wish to produce estimates. Usually, a cell is defined by a time period, an age range, and a sex. So, for example, a cell might be women who were age 30-34 in 2015.

We'll calculate estimates from the sibling histories in three main steps:

1. Create an **esc** dataset, so called because there is  row for each **e**go X **s**ibling X **c**ell
2. Aggregate this esc dataset up into an **ec** dataset, which has a row for the reports made by each ego for each cell
3. Aggregate this ec dataset up into estimates for death rates, using either the individual or aggregate visibility approach (or both)

### Laying the groundwork

First, we'll need to

* add an indicator for frame population membership
* create an object that describes the cells we wish to generate reports for

TODO DESCRIBE

```{r}
ex.sib <- ex.sib %>% 
  mutate(in.F = as.numeric((alive==1) & (age >= 15) & (age <= 49) & (sex == 'f')))
```

TODO DESCRIBE

```{r}
with(ex.sib, table(in.F, useNA='ifany'))
```

These sibs with missing `in.F` turn out not to have ages recorded, Since we 
need to be able to determine whether or not each sibling is on in the frame population, we will
drop them.

```{r}
ex.sib <- ex.sib %>% filter(! is.na(in.F))
``


TODO DESCRIBE

```{r}
cc <- cell_config(age.groups='5yr', 
                  time.periods='5yr_beforeinterview',
                  start.obs='dob',
                  end.obs='endobs',
                  event='death.date',
                  age.offset='dob',
                  time.offset='doi',
                  exp.scale=1/12)
```



### Estimating death rates

To make these three steps reasonably straightforward, we've written a function, `sibling_estimator`, that takes care of all three of them.

```{r}
ex_ests <- sibling_estimator(sib.dat = ex.sib %>% mutate(const_weight = 1, sib.sex=sex),
                             ego.id = 'caseid',
                             sib.id = 'sibid',
                             sib.frame.indicator = 'in.F',
                             sib.sex = 'sib.sex',
                             cell.config=cc,
                             weights='wwgt')

ex_ests$asdr.ind
```

```{r}
ggplot(ex_ests$asdr.ind) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, group=sib.sex)) +
  theme_minimal() +
  scale_y_log10() + 
  ggtitle('individual estimator')
```



```{r}
ggplot(ex_ests$asdr.agg) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, group=sib.sex)) +
  theme_minimal() +
  scale_y_log10() + 
  ggtitle('aggregate estimator')
```

```{r}
compare <- bind_rows(ex_ests$asdr.ind, ex_ests$asdr.agg)

ggplot(compare) +
  geom_line(aes(x=sib.age, y=1000*asdr.hat, color=sib.sex, linetype=estimator, group=interaction(estimator, sib.sex))) +
  theme_minimal() +
  #facet_grid(sib.sex ~ .) +
  facet_grid(. ~ sib.sex) +
  scale_y_log10() + 
  ggtitle('both estimators')

```

